---
title: "Case Studies: Cleaning Data"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
# Load packages
library(dplyr)
library(naniar)
```

In the following document we pre-process the case study data.
```{r}
# Load in all case study data
adult_train = read.csv("adult/adult.data", header = FALSE) # census income (adult)
adult_test = read.csv("adult/adult.test", header = FALSE, skip = 1)
diabetes = read.csv("diabetes/diabetic_data.csv") # diabetes
```


# Census Income (Adult 50k) data
## Pre-processing
```{r}
# Merge adult train and adult test together to create one large dataset
adult = rbind(adult_train, adult_test)
```

```{r}
dim(adult)
```

```{r}
# Assign variable names to the data
colnames(adult) = c("age", "workclass", "fnlwgt", "education", "education-num", 
                    "marital-status", "occupation", "relationship", "race", 
                    "sex", "capital-gain", "capital-loss", "hours-per-week", 
                    "native-country", "income")
```

We check the number of missing values
```{r}
# Missing values are coded as ? instead of NA
# Recode to NA
adult[adult == " ?"] = NA
sum(is.na(adult))
```

The number of missing values is 6465.

Next, we check what the percentage of missing values is
```{r}
pct_miss(adult)
```

We use a complete case analysis, as the percentage is lower than 5%.
```{r}
# Make complete cases
adult = adult %>%
  filter(complete.cases(.))
```

We remove not needed variables and clean the data
```{r}
# Remove variables that are not needed (fnlwgt) or are a replication of another variable (education-num)
adult = select(adult, -fnlwgt, -'education-num')

# Transform age from character to numeric
adult$age = as.numeric(adult$age)
```

```{r}
# Check the outcome variable, which shows outcomes are coded both with and without .
table(adult$income)

# Remove whitespace form start and end of string for binary variables (if present)
adult$income = gsub("^\\s+|\\s+$", "", adult$income) 
adult$sex = gsub("^\\s+|\\s+$", "", adult$sex)

# remove . from test data at end string
adult$income = gsub("\\.$", "", adult$income) 

table(adult$income)
```

Next, we transform variables into factors and rename some levels
```{r}
# Define binary factors 
binary_factor = c("income", "sex")
adult[, binary_factor]= lapply(adult[, binary_factor], as.factor)
adult$income = as.numeric(adult$income == ">50K") # change to 0-1 (1 = >50K)
adult$sex = as.numeric(adult$sex == "Male") # Change to 0-1 (Male = 1)

# Define factors
nominal_factor = c("workclass", "education", "marital-status", "occupation", 
                   "relationship", "race", "native-country")

# Remove whitespace (often a space is before each entry) and transform into a factor
for (col in nominal_factor) {
  adult[[col]] <- gsub("^\\s+|\\s+$", "", adult[[col]])
  adult[[col]] = as.factor(adult[[col]])
}
str(adult)

# Rename columns
adult = adult %>%
  rename(
    marital_status = `marital-status`,
    native_country = `native-country`,
    capital_gain = `capital-gain`,
    capital_loss = `capital-loss`,
    hours_per_week = `hours-per-week`
  )

# Change outcome variable to y
colnames(adult)[13] = "y"
```


A summary
```{r}
summary(adult)
```

Finally, we save the cleaned data to be used in our analysis.
```{r}
write.csv(x = adult, file = "adult_cleaned.csv", row.names = FALSE)
```



# Diabetes data
## Pre-processing
We check the number of missing values
```{r}
# Missing values are coded as ? instead of NA
# Recode to NA
diabetes[diabetes == "?"] = NA

# Check number of missing values
sum(is.na(diabetes))
```

The number of missing values is 192849.

# Removal of variables
weight, payer_code, and medical_specialty have many missing values. The authors removed weight and payer_code, but kept medical_specialty. We remove all three.
```{r}
colSums(is.na(diabetes))
```
```{r}
# Remove the covariates with many missing values
diabetes = diabetes %>% select(-weight, -payer_code, -medical_specialty)
```

Next, we check what the percentage of missing values is
```{r}
pct_miss(diabetes)
```

We use a complete case analysis, as the percentage is lower than 5%.
```{r}
diabetes = diabetes %>%
  filter(complete.cases(.))
```

### Outcome variable readmitted
The researchers were interested in early readmission. So, readmitted within 30 days is class 1, while readmitted after 30 days or no readmission are of class 0.
```{r}
diabetes$y <- ifelse(diabetes$readmitted == "<30", 1, 0)
```

### Visit
Researchers also only considered the first encounter of the patient.
```{r}
diabetes = diabetes %>%
  group_by(patient_nbr) %>%  
  slice(1) %>%  # select only the first entry of each patient
  ungroup()
```

The original authors also deleted patients who died or were admitted to a hospice. We did not to this, as this was not very clearly specified in their paper and no clear steps were provided.


### Recoding
`diag_1`, `diag_2`, and `diag_3` have 692, 716, and 756 distinct categories that correspond to diagnoses, respectively. That are a lot of factors. By consulting the paper for this data, we use Table 2 to group the categories into 9 more general ones. The categories correspond to 9 groups. Table 2 shows the icd9 codes that correspond to the groups.

```{r}
for (var in c("diag_1", "diag_2", "diag_3")) {
  
  diabetes[[paste0(var, "_group")]] = "Other"
  
  diabetes[[paste0(var, "_group")]][diabetes[[var]] %in% 390:459 | diabetes[[var]] == 785] = "Circulatory"
  
  diabetes[[paste0(var, "_group")]][diabetes[[var]] %in% 460:519 | diabetes[[var]] == 786] = "Respiratory"
  
  diabetes[[paste0(var, "_group")]][diabetes[[var]] %in% 520:579 | diabetes[[var]] == 787] = "Digestive"
  
  diabetes[[paste0(var, "_group")]][grepl("^250\\.", diabetes[[var]])] = "Diabetes"
  
  diabetes[[paste0(var, "_group")]][diabetes[[var]] %in% 800:999] = "Injury"
  
  diabetes[[paste0(var, "_group")]][diabetes[[var]] %in% 710:739] = "Musculoskeletal"
  
  diabetes[[paste0(var, "_group")]][diabetes[[var]] %in% 580:629 | diabetes[[var]] == 788] = "Genitourinary"
  
  diabetes[[paste0(var, "_group")]][diabetes[[var]] %in% 140:239] = "Neoplasms"
  
}

```


### Cleaning and Recoding
We remove variables that are not important predictors as they have been recoded or are only identifiers.
```{r}
diabetes = diabetes %>% select(-encounter_id, -patient_nbr, -diag_1, -diag_2, -diag_3, -readmitted)
```


```{r}
# Name all variables that are factors
factors_variables = c(
  "race", "gender", "age", "admission_type_id", "discharge_disposition_id", 
  "admission_source_id", "max_glu_serum", "A1Cresult", "metformin", "repaglinide", 
  "nateglinide", "chlorpropamide", "glimepiride", "acetohexamide", "glipizide", 
  "glyburide", "tolbutamide", "pioglitazone", "rosiglitazone", "acarbose", 
  "miglitol", "troglitazone", "tolazamide", "examide", "citoglipton", "insulin", 
  "glyburide.metformin", "glipizide.metformin", "glimepiride.pioglitazone", 
  "metformin.rosiglitazone", "metformin.pioglitazone", "change", "diabetesMed", 
  "y", "diag_1_group", "diag_2_group", "diag_3_group"
  )

# Recode to class factor
diabetes[, factors_variables]= lapply(diabetes[, factors_variables], as.factor)
```


We check the summary
```{r}
summary(diabetes)
```

examide, citoglipton, glimepiride.pioglitazone, metformin.rosiglitazone only have entry No. Therefore, we remove variables with only 1 level
```{r}
diabetes = diabetes %>% select(-examide, -citoglipton, -`glimepiride.pioglitazone`, -`metformin.rosiglitazone`)
```

```{r}
dim(diabetes)
```

Finally, we save the cleaned data to be used in our analysis.
```{r}
write.csv(x = diabetes, file = "diabetes_cleaned.csv", row.names = FALSE)
```








